Human capital mediates natural selection in contemporary humans

### 2021-12-10 why-natural-selection.Rmd
Reorder income table; discuss in text

n_increases <- sum(res_ineq$ratio > 1)
median_pct_change <- (median(res_ineq$ratio) - 1) * 100
```

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases the correlation of polygenic scores with income 
for `r n_increases` out of 33 polygenic
scores, with a median increase in the correlation of `r median_pct_change`% 
(Appendix Table \@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
important for understanding class mobility and structure [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantially
important. Since many polygenic scores are predictive of disease risk, they
could also potentially increase health inequalities. In general, the
evolutionary history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may aid in mapping the genetic architecture of current health disparities.

@@

Table \@ref(tab:tbl-inequality-effects) shows correlations between children's
polygenic scores and household income (UKB data field 738). Column "With
selection" uses respondents' scores weighted by Age/Qualification, times number
of children. Column "Without selection" uses scores weighted by
Age/Qualification only, i.e. if all respondents had the same number of children,
equivalently the correlation among respondents.

@@

res_ineq %>% 
      select(
        PGS                      = score, 
        `Cor. without selection` =cf,
        `Cor. with selection`    = actual, 
        `Ratio`                  = ratio
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`Ratio`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, 2:3, 3) %>% 
      set_number_format(everywhere, 4, 2) %>% 
      set_caption("Correlations of polygenic scores with income group.") %>% 
      theme_compact() %>% 
      set_width(0.75) %>% 
      set_col_width(c(.4, .2, .2, .2)) %>% 
      set_bottom_border(final(1), everywhere) %>% 
      set_position("center")
```

>>>>

n_increases <- sum(res_ineq$ratio > 1)
median_pct_change <- (median(res_ineq$ratio) - 1) * 100

cor_ea3 <- res_ineq[res_ineq$score == "EA3_excl_23andMe_UK", ]
cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100
```

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
proportionate increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
important for understanding class mobility and structure [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. Since
many polygenic scores are predictive of disease risk, they could also
potentially increase health inequalities. In general, the evolutionary history
of anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

Table \@ref(tab:tbl-inequality-effects) shows correlations between children's
polygenic scores and household income (UKB data field 738). Column "With
selection" uses respondents' scores weighted by Age/Qualification, times number
of children. Column "Without selection" uses scores weighted by
Age/Qualification only, i.e. the counterfactual correlation if all respondents
had the same number of children (equivalently, the correlation in the
respondents' own generation).

@@

res_ineq %>% 
      select(
        PGS                      = score, 
        `Cor. with selection`    = actual, 
        `Cor. without selection` = cf,
        `Ratio`                  = ratio
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`Cor. with selection`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, 2:3, 3) %>% 
      set_number_format(everywhere, 4, 2) %>% 
      set_caption("Correlations of polygenic scores with income group.") %>% 
      theme_compact() %>% 
      set_width(0.75) %>% 
      set_col_width(c(.4, .2, .2, .2)) %>% 
      set_bottom_border(final(1), everywhere) %>% 
      set_position("center")
```
